Approximation as Differentiation

### 2025-05-08 main/formal-structures.tex
Progress on properties of LatGal.

First we introduce the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward Galois slicing. Galois connections are pairs of monotone functions between
posets, where each is the closest approximation (one from above, one from below) to an inverse of the other.

\begin{definition}[Galois connection]
Suppose $X$ and $Y$ are posets. A \emph{Galois connection} $f \adj g: X \to Y$ is a pair of monotone functions
$f: X \to Y$ and $g: Y \to X$ satisfying $y \leq f(x) \iff g(y) \leq x$ for any $x \in X$ and $y \in Y$.
\end{definition}

\noindent The $\adj$ notation is justified because a Galois connection $f \adj g: X \to Y$ can also be seen an
adjunction between poset categories, with monotone $f$ and $g$ interpreted as functors; $f$ is usually
referred to as the \emph{upper} (right) adjoint and $g$ as the \emph{lower} (left) adjoint. Galois connections
compose component-wise, with $\id_X \adj \id_X: X \to X$ as the unit for composition, and thus form a category
where the objects are posets and the morphisms are Galois connections.

As shown in \secref{introduction:galois-slicing}, for Galois slicing we require a setting where the
approximants of a point form a bounded lattice, with least element $\bot$ representing the approximation that
discards all information and greatest element $\top$ representing the approximation that discards no
information. So we define the following subcategory.

\begin{definition}
Define the category $\LatGal$ which has as objects $X, Y$ all bounded lattices, and as morphisms all Galois
connections $f \adj g: X \to Y$.
\end{definition}

\noindent Right adjoints preserves limits and left adjoints preserves colimits, so for any $f \adj g: X \to Y$
in $\LatGal$, $f$ and $g$ preserve the meet-semilattice structure of $X$ and join-semilattice structure of $Y$
respectively.

@@

The product/coproduct is called a \emph{biproduct}, and the biproduct structure is denoted by $(\biprod, 0)$.
The unit $0$ is both terminal and initial and is called a \emph{zero} object. A semi-additive category
$\cat{C}$ is enriched in $\CMon$, the category of commutative monoids: for any two morphisms $f, g: X \to Y$
in $\cat{C}$, the biproduct structure provides a way to ``add'' them together, forming a morphism $f + g: X
\to Y$. Diagrammatically:

>>>>

First we introduce the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward Galois slicing. Galois connections are pairs of monotone functions $f: Y \to
X$ and $g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse
of $g$, and $g$ the best approximation from above to an inverse of $f$.

\begin{definition}[Galois connection]
Suppose $X$ and $Y$ are posets. A \emph{Galois connection} $f \adj g: X \to Y$ is a pair of monotone functions
$f: Y \to X$ and $g: X \to Y$ satisfying $y \leq f(x) \iff g(y) \leq x$ for any $x \in X$ and $y \in Y$.
\end{definition}

\noindent Galois connections thus generalise order isomorphisms. The $\adj$ notation is justified because a
Galois connection $f \adj g: X \to Y$ can also be seen an adjunction between poset categories, with monotone
$f$ and $g$ interpreted as functors; $f$ is usually referred to as the \emph{upper} (right) adjoint and $g$ as
the \emph{lower} (left) adjoint. Galois connections compose component-wise, with $\id_X \adj \id_X: X \to X$
as the unit for composition, and thus form a category $\PosGal$ with all posets as objects and all Galois
connections between them as morphisms.

As sketched in \secref{introduction:galois-slicing}, for Galois slicing we require a setting where the
approximants of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$
representing the approximation that discards all information about $x$, greatest element $\top$ representing
the approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways
to combine approximations. Thus rather than working in $\PosGal$, we define the following subcategory.

\begin{definition}
Define the category $\LatGal$ which has as objects $X = (X, \meet, \join, \top, \bot)$ all bounded lattices,
and as morphisms all Galois connections $f \adj g: X \to Y$.
\end{definition}

\noindent Right adjoints preserves limits and left adjoints preserves colimits, so for any $f \adj g: X \to Y$
in $\LatGal$, $f$ is a \emph{meet-semilattice homomorphism}, i.e.~preserves the meet-semilattice structure
$(X, \meet, \top)$. Similarly, $g$ is a join-semilattice homomorphism with respect to $(X, \join, \bot)$.

\subsubsection{Additive structure on morphisms.}

A bounded semilattice is an example of a \emph{commutative monoid} $(X, +, 0)$, namely a set $X$ equipped with
an associative, commutative binary operation $+$ with unit $0$. One observation about commutative monoids is
that pointwise lifting of $+$ to homomorphisms $f, g: X \to Y$, where $(f + g)(x) = f(x) + g(x): X \to Y$, is
also ``additive'', in the sense of being associative and commutative and having a ``zero'' homomorphism
$0_{X,Y}: X \to Y$ given by the constant map sending any $x \in X$ to $0_Y$, which acts as a unit for addition
of homomorphisms. The homomorphisms from $X$ to $Y$ thus have a commutative monoid structure themselves.

In particular $\LatGal$ has this structure..

% Moreover since $(X, \top, \meet)$ is a commutative monoid, and thus a $\CMon$-enriched (poset)
% category, for any $X, Y$ there is a zero meet-semilattice homomorphism $0^{\meet}_{X,Y} = \const(\top_Y): X
% \to Y$ and also (idempotent) addition of homomorphisms $f +_{\meet} g: X \to Y$ given pointwise by $\meet$.
% Dually, there is a zero join-semilattice homomorphism $0^{\join}_{X,Y} = \const(\bot_Y): X \to Y$ and
% idempotent addition of homomorphisms $f +_{\join} g: X \to Y$ given pointwise by $\join$.

@@

The product/coproduct is called a \emph{biproduct}, and the biproduct structure is denoted by $(\biprod, 0)$.
The unit $0$ is both terminal and initial and is called a \emph{zero} object. \todo{State biproduct ``laws''
that follow from this formulation.} A semi-additive category $\cat{C}$ is enriched in $\CMon$, the category of
commutative monoids: for any two morphisms $f, g: X \to Y$ in $\cat{C}$, the biproduct structure provides a
way to ``add'' them together, forming a morphism $f + g: X \to Y$. Diagrammatically:
